---
title: "Vinha et al. 2024, Ensemble modelling to predict the distribution of vulnerable marine ecosystems indicator taxa on data-limited seamounts of Cabo Verde (NW Africa) - Calculate terrain variables from MB bathymetry and mean oceanographical variables from Viking20X model"
author: "Beatriz Vinha"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
library(rgdal)
library(raster)
library(terra)
library(spatialEco)
library(sf)
library(MultiscaleDTM) #from https://github.com/ailich/MultiscaleDTM
library(ncdf4) 
library(rgdal)
```

Here, the environmental data used
for modelling is calculated and prepared:

1.  Using a multi-scale approach, we calculate different terrain
    derivatives from a Multibeam bathymetry grid, representing all data
    available for the Cabo Verde archipelago, at 100
    m of resolution.

2.  Extract data from the VIKING20X oceanographic model data (Biastoch
    et al., 2021), in order to calculate the mean value of each variable
    and to extract it as a geo-referenced raster. Data from the
    VIKING20X model corresponds to bottom values of temperature,
    salinity and ocean velocities (U and V) for the Cabo Verde region.
    The temporal resolution of the data is from December 2009 to
    December 2019.

**Data sources:**

-   Bathymetry data and model mask can be downloaded from: 

Vinha, B., Murillo, F. J., Schumacher, M., Hansteen, T., Schwarzkopf, F., Biastoch, A., 
Kenchington, E., Piraino, S., Orejas, C., Huvenne, V. (2024). Terrain variables used for 
ensemble distribution modelling of vulnerable marine ecosystems indicator taxa on data-limited 
seamounts of Cabo Verde (NW Africa) [Dataset]. Dryad. https://doi.org/10.5061/dryad.0vt4b8h5g

-   Oceanographical data can be downloaded from:

Schwarzkopf, F. (2024). Supplementary data to Vinha et al. (2024): Ensemble modelling to 
predict the distribution of Vulnerable Marine Ecosystems indicator taxa on data-limited 
seamounts of Cabo Verde (NW Africa) [dataset]. GEOMAR Helmholtz Centre for Ocean Research 
Kiel [distributor]. hdl:20.500.12085/20248b0a-49fd-4868-90bf-581c61f4b396

**Reference to the study where data and code was used**

Vinha,B., Murillo,F. J.,Schumacher,M., Hansteen,T. H.,Schwarzkopf,F. U.,
Biastoch,A., Kenchington,E.,Piraino,S.,Orejas,C., & Huvenne,V. A. I.(2024).
Ensemble modelling to predict the distribution of vulnerable marine ecosystems indicator 
taxa on data-limited seamounts of Cabo Verde (NW Africa).Diversity and Distributions,
00, e13896.https://doi.org/10.1111/ddi.13896

# 1. Load data

```{r}
#Import bathymetry file
bathy_cv <- raster("00_Caboverde_bathy_100m_UTM26N.tif") 
plot(bathy_cv)
```

```{r}
##Load mask
seamount_mask <- shapefile("Seamounts_mask_poly.shp")
bathyseamount_m <- mask(bathy_cv, seamount_mask)
```

```{r}
#Import oceanographical data (in nc file format)
btsal_data <- nc_open('1_VIKING20X.L46-KFS003_1m_20090101_20191231_T_S_CapeVerde_bottom.nc')
u_data <- nc_open('1_VIKING20X.L46-KFS003_1m_20090101_20191231_U_CapeVerde_bottom.nc')
v_data <- nc_open('1_VIKING20X.L46-KFS003_1m_20090101_20191231_V_CapeVerde_bottom.nc')
```

# 2. Calculate Terrain Variables from MB Bathymetry 

## a. Calculate Slope

```{r}
Slope_3x3 <- SlpAsp(bathyseamount_m, w = c(3,3), unit = "degrees", method = "queen", metrics = c("slope"))
Slope_9x9 <- SlpAsp(bathyseamount_m, w = c(9,9), unit = "degrees", method = "queen", metrics = c("slope"))
Slope_17x17 <- SlpAsp(bathyseamount_m, w = c(17,17), unit = "degrees", method = "queen", metrics = c("slope"))
Slope_33x33 <- SlpAsp(bathyseamount_m, w = c(33,33), unit = "degrees", method = "queen", metrics = c("slope"))

par(mfrow=c(2,2))
plot(Slope_3x3, main = "Slope 3x3")
plot(Slope_9x9, main = "Slope 9x9")
plot(Slope_17x17, main = "Slope 17x17")
plot(Slope_33x33, main = "Slope 33x33")

#Download Rasters
writeRaster(Slope_3x3, "01_Slope_3x3", format = "GTiff", overwrite = TRUE)
writeRaster(Slope_9x9, "01_Slope_9x9", format = "GTiff", overwrite = TRUE)
writeRaster(Slope_17x17, "01_Slope_17x17", format = "GTiff", overwrite = TRUE)
writeRaster(Slope_33x33, "01_Slope_33x33", format = "GTiff", overwrite = TRUE)
```

## b. Calculate Aspect (Eastness and Northness)

```{r}
#####EASTNESS#####
eastness_3x3 <- SlpAsp(bathyseamount_m, w = c(3,3), unit = "radians", method = "queen", metrics = c("eastness"))
eastness_9x9 <- SlpAsp(bathyseamount_m, w = c(9,9), unit = "radians", method = "queen", metrics = c("eastness"))
eastness_17x17 <- SlpAsp(bathyseamount_m, w = c(17,17), unit = "radians", method = "queen", metrics = c("eastness"))
eastness_33x33 <- SlpAsp(bathyseamount_m, w = c(33,33), unit = "radians", method = "queen", metrics = c("eastness"))

par(mfrow=c(2,2))
plot(eastness_3x3, main = "Eastness 9x9" )
plot(eastness_9x9, main = "Eastness 9x9")
plot(eastness_17x17, main = "Eastness 17x17")
plot(eastness_33x33, main = "Eastness 33x33")

#Download Rasters
writeRaster(eastness_3x3, "02_eastness_3x3", format = "GTiff", overwrite = TRUE)
writeRaster(eastness_9x9, "02_eastness_9x9", format = "GTiff", overwrite = TRUE)
writeRaster(eastness_17x17, "02_eastness_17x17", format = "GTiff", overwrite = TRUE)
writeRaster(eastness_33x33, "02_eastness_33x33", format = "GTiff", overwrite = TRUE)

#####NORTHNESS#####
northness_3x3 <- SlpAsp(bathyseamount_m, w = c(3,3), unit = "radians", method = "queen", metrics = c("northness"))
northness_9x9 <- SlpAsp(bathyseamount_m, w = c(9,9), unit = "radians", method = "queen", metrics = c("northness"))
northness_17x17 <- SlpAsp(bathyseamount_m, w = c(17,17), unit = "radians", method = "queen", metrics = c("northness"))
northness_33x33 <- SlpAsp(bathyseamount_m, w = c(33,33), unit = "radians", method = "queen", metrics = c("northness"))

par(mfrow=c(2,2))
plot(northness_3x3, main = "northness 9x9" )
plot(northness_9x9, main = "northness 9x9")
plot(northness_17x17, main = "northness 17x17")
plot(northness_33x33, main = "northness 33x33")

#Download Rasters
writeRaster(northness_3x3, "03_northness_3x3", format = "GTiff", overwrite = TRUE)
writeRaster(northness_9x9, "03_northness_9x9", format = "GTiff", overwrite = TRUE)
writeRaster(northness_17x17, "03_northness_17x17", format = "GTiff", overwrite = TRUE)
writeRaster(northness_33x33, "03_northness_33x33", format = "GTiff", overwrite = TRUE)
```

## c. Calculate Terrain Curvature

```{r}
qmetrics_3x3 <- Qfit(bathyseamount_m, w = c(3,3), unit = "degrees", na.rm = TRUE)
qmetrics_9x9 <- Qfit(bathyseamount_m, w = c(9,9), unit = "degrees", na.rm = TRUE)
qmetrics_17x17 <- Qfit(bathyseamount_m, w = c(17,17), unit = "degrees", na.rm = TRUE)
qmetrics_33x33 <- Qfit(bathyseamount_m, w = c(33,33), unit = "degrees", na.rm = TRUE)

#####Profile curvature#####
profc_3x3 <- subset(qmetrics_3x3, "profc")
profc_9x9 <- subset(qmetrics_9x9, "profc")
profc_17x17 <- subset(qmetrics_17x17, "profc")
profc_33x33 <- subset(qmetrics_33x33, "profc")

par(mfrow=c(2,2))
plot(profc_3x3, main="Profile Curvature 3x3")
plot(profc_9x9, main="Profile Curvature 9x9")
plot(profc_17x17, main="Profile Curvature 17x17")
plot(profc_33x33, main="Profile Curvature 33x33")

writeRaster(profc_3x3, "04_profc_3x3", format = "GTiff", overwrite = TRUE)
writeRaster(profc_9x9, "04_profc_9x9", format = "GTiff", overwrite = TRUE)
writeRaster(profc_17x17, "04_profc_17x17", format = "GTiff", overwrite = TRUE)
writeRaster(profc_33x33, "04_profc_33x33", format = "GTiff", overwrite = TRUE)

#####Planar curvature#####
planc_3x3 <- subset(qmetrics_3x3, "planc")
planc_9x9 <- subset(qmetrics_9x9, "planc")
planc_17x17 <- subset(qmetrics_17x17, "planc")
planc_33x33 <- subset(qmetrics_33x33, "planc")

par(mfrow=c(2,2))
plot(planc_3x3, main="Plan Curvature 3x3")
plot(planc_9x9, main="Plan Curvature 9x9")
plot(planc_17x17, main="Plan Curvature 17x17")
plot(planc_33x33, main="Plan Curvature 33x33")

writeRaster(planc_3x3, "05_planc_3x3", format = "GTiff", overwrite = TRUE)
writeRaster(planc_9x9, "05_planc_9x9", format = "GTiff", overwrite = TRUE)
writeRaster(planc_17x17, "05_planc_17x17", format = "GTiff", overwrite = TRUE)
writeRaster(planc_33x33, "05_planc_33x33", format = "GTiff", overwrite = TRUE)

#####Mean curvature#####
meanc_3x3 <- subset(qmetrics_3x3, "meanc")
meanc_9x9 <- subset(qmetrics_9x9, "meanc")
meanc_17x17 <- subset(qmetrics_17x17, "meanc")
meanc_33x33 <- subset(qmetrics_33x33, "meanc")

par(mfrow=c(2,2))
plot(meanc_3x3, main="Mean Curvature 3x3")
plot(meanc_9x9, main="Mean Curvature 9x9")
plot(meanc_17x17, main="Mean Curvature 17x17")
plot(meanc_33x33, main="Mean Curvature 33x33")

#Download Rasters
writeRaster(meanc_3x3, "06_meanc_3x3", format = "GTiff", overwrite = TRUE)
writeRaster(meanc_9x9, "06_meanc_9x9", format = "GTiff", overwrite = TRUE)
writeRaster(meanc_17x17, "06_meanc_17x17", format = "GTiff", overwrite = TRUE)
writeRaster(meanc_33x33, "06_meanc_33x33", format = "GTiff", overwrite = TRUE)
```

## d. Calculate Topographic Position Index (TPI)

```{r}
######Fine TPI#####
ftpi_300m <- tpi(bathyseamount_m, win = "rectangle", scale = 3) 

######Broad TPI######
btpi_1500m <- tpi(bathyseamount_m, win = "rectangle", scale = 15) 

#Plot
par(mfrow=c(1,2))
plot(ftpi_300m, main="Fine TPI 300")
plot(btpi_1500m, main="Broad TPI 1500")

#Download Rasters
writeRaster(ftpi_300m, "07_ftpi_300m", format = "GTiff", overwrite = TRUE)
writeRaster(btpi_1500m, "07_btpi_1500m", format = "GTiff", overwrite = TRUE)
```

## e. Calculate Roughness

```{r}
rough <- terrain(bathyseamount_m, opt="roughness")
plot(rough, main="Roughness")

#Download Rasters
writeRaster(rough, "08_roughness", format = "GTiff")
```

## f. Calculate Terrain Ruggedness Index (TRI)

```{r}
tri_3x3 <- terrain(bathyseamount_m, opt=c('tri'))
plot(tri_3x3, main="TRI (Ruggedness)")

#Download Rasters
writeRaster(tri_3x3, "09_tri_3x3", format = "GTiff")
```

## g. Calculate Vector Ruggedness Measure (VRM)

```{r}
vrm_3x3 <- vrm(bathyseamount_m, s=3)
vrm_33x33 <- vrm(bathyseamount_m, s=33)

par(mfrow=c(1,2))
plot(vrm_3x3, main="Vector Ruggedness Measure 3x3")
plot(vrm_33x33, main="Vector Ruggedness Measure 33x33")

writeRaster(vrm_3x3, "10_vrm_3x3", format = "GTiff", overwrite = TRUE)
writeRaster(vrm_33x33, "10_vrm_33x33", format = "GTiff", overwrite = TRUE)
```

# 3. Oceanographic data

Below, the mean value of each oceanographic variable is calculated and
the final output is converted to a raster.

## a. Bottom Temperature and Salinity

```{r}
#extract lat, lon and variable (i.e. bottomT and Sali)
lon <- ncvar_get(btsal_data, "nav_lon")
lat <- ncvar_get(btsal_data, "nav_lat", verbose = F)
bt <- ncvar_get(btsal_data, "votemper")
sal <- ncvar_get(btsal_data, "vosaline")
```

Each variable will be converted to an array with 3 dimensions (lon, lat
and time) and by "slicing" it you will select the layer correspondent to
the first n time (i.e. month) of the data set (n layer); e.g "bT.slice1
\<- bt[, , 1]", slices the first month of the dataset.

However, I am interested in the mean value of the variable for the whole
dataset, therefore, I will calculate the mean of the third dimension.

```{r}
bt.mean <- apply(bt, c(1,2), mean)
sal.mean <- apply(sal, c(1,2), mean)
```

Now, create a raster:

```{r}
#Create raster with mean bottom temperature
bt.r <- raster(t(bt.mean), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat), crs=CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs+ towgs84=0,0,0"))


#flip raster
bt.r <- flip(bt.r, direction='y')
plot(bt.r)
```

```{r}
#Create raster with mean bottom salinity
sal.r <- raster(t(sal.mean), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat), crs=CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs+ towgs84=0,0,0"))


#flip raster
sal.r <- flip(sal.r, direction='y')
plot(sal.r)
```

## b. Current Magnitude and Speed

Now, repeat for U and V vectors.

```{r}
######U CURRENT
#extract lat, lon and variable 
lon1 <- ncvar_get(u_data, "nav_lon")
lat1 <- ncvar_get(u_data, "nav_lat", verbose = F)
u <- ncvar_get(u_data, "vozocrtx")

#Calculate the mean of the third dimension
u.mean <- apply(u, c(1,2), mean)

#Convert to raster
u.r <- raster(t(u.mean), xmn=min(lon1), xmx=max(lon1), ymn=min(lat1), ymx=max(lat1), crs=CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs+ towgs84=0,0,0"))


#flip raster
u.r <- flip(u.r, direction='y')
plot(u.r)
```

```{r}
######V CURRENT
#extract lat, lon and variable 
lon2 <- ncvar_get(v_data, "nav_lon")
lat2 <- ncvar_get(v_data, "nav_lat", verbose = F)
v <- ncvar_get(v_data, "vomecrty")

#Calculate the mean of the third dimension
v.mean <- apply(v, c(1,2), mean)

#Convert to raster
v.r <- raster(t(v.mean), xmn=min(lon2), xmx=max(lon2), ymn=min(lat2), ymx=max(lat2), crs=CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs+ towgs84=0,0,0"))


#flip raster
v.r <- flip(v.r, direction='y')
plot(v.r)
```

And, calculate mean bottom current speed.

```{r}
#Apply formula to calculate current speed
cs <- sqrt((v^2)+(u^2))

#calculate mean
cs.mean <- apply(cs, c(1,2), mean)

#Create raster
cs.r <- raster(t(cs.mean), xmn=min(lon2), xmx=max(lon2), ymn=min(lat2), ymx=max(lat2), 
               crs=CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs+ towgs84=0,0,0"))
#flip raster
cs.r <- flip(cs.r, direction='y')

#Plot
plot(cs.r)
```

## c. Re-project layers to UTM26N and save as raster files

```{r}
#Reproject
crs <- "+proj=utm +zone=26 +ellps=WGS72 +units=m +no_defs"

meanSal_utm26 <- projectRaster(sal.r, crs=crs)
meanbT_utm26 <- projectRaster(bt.r, crs=crs)
meanU_utm26 <- projectRaster(u.r, crs=crs)
meanV_utm26 <- projectRaster(v.r, crs=crs)
meanCS_utm26 <- projectRaster(cs.r, crs=crs)
```

```{r}
#Plot
par(mfrow=c(2,2))
plot(meanSal_utm26, main = "Salinity utm26")
plot(meanbT_utm26, main = "Temperature utm26")
plot(meanU_utm26, main = "U utm26")
plot(meanV_utm26, main = "V utm26")
```

```{r}
#Save
writeRaster(meanSal_utm26, "meanSal_utm26.tif", "GTiff", overwrite=TRUE)
writeRaster(meanbT_utm26, "meanbT_utm26.tif", "GTiff", overwrite=TRUE)
writeRaster(meanU_utm26, "meanU_utm26.tif", "GTiff", overwrite=TRUE)
writeRaster(meanV_utm26, "meanV_utm26.tif", "GTiff", overwrite=TRUE)
```
